BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-11-13 ms.tex

However, in practice it is challenging to optimize adiabatic EMRI waveform calculations in order to make them efficient and accessible, so that they can be incorporated in large scale samplings of parameter space. Currently, only one open-source tool exists for producing adiabatic EMRI waveforms: the Fast EMRI Waveforms (FEW) Python package \cite{ChuaGallVall19, ChuaETC20, KatzETC20, KatzETC21}. While this tool is the gold standard for EMRI waveform models, particularly due to its ability to leverage GPUs, it is currently restricted to eccentric binaries with non-rotating black holes. Since we expect almost all astrophysical EMRIs to possess a rotating massive black hole, it is important to extend models such as FEW into this crucial area of parameter space.

@@

It is important to note that \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a variety of massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin (which is a post-adiabatic effect) \cite{SkouLuke22, DrumETC23}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

@@

\begin{acknowledgements}
This research was supported by an appointment to the NASA Postdoctoral Program at the NASA Goddard Space Flight Center, administered by Oak Ridge Associated Universities under contract with NASA. The author thanks Michael Katz and Lorenzo Speri for extensive discussions and advice on EMRI data analysis, and also thanks John G.~Baker for their insight and helpful feedback. This work makes use of the Black Hole Perturbation Toolkit. It also uses the \texttt{Eryn}\footnote{\href{https://github.com/mikekatz04/Eryn}{https://github.com/mikekatz04/Eryn}} and \texttt{lisatools}\footnote{\href{https://github.com/mikekatz04/LISAanalysistools}{https://github.com/mikekatz04/LISAanalysistools}} repositories authored by Michael Katz.
\end{acknowledgements}

>>>>

However, in practice it is challenging to optimize adiabatic EMRI waveform calculations in order to make them efficient and accessible, so that they can be incorporated in large scale samplings of parameter space. Currently, there is one main open-source self-force model for producing adiabatic EMRI waveforms: the Fast EMRI Waveforms (FEW) Python package \cite{ChuaGallVall19, ChuaETC20, KatzETC20, KatzETC21}. While this tool is the gold standard for EMRI waveform models and data analysis, particularly due to its ability to leverage GPUs, it is currently restricted to eccentric binaries with non-rotating black holes. Since we expect almost all astrophysical EMRIs to possess a rotating massive black hole, it is important to extend models such as FEW into this crucial area of parameter space.

@@

Importantly, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}, and other works have even modeled eccentric, equatorial and eccentric, inclined (precessing) EMRIs for a variety of massive black hole spins \cite{FujiShib20, SkouLuke22, HughETC21}. Recent works have also produced waveforms that include the effects due to the small body's spin (which is a post-adiabatic effect) \cite{SkouLuke22, DrumETC23}, while effective-one-body models have combined post-Newtonian and GSF information to provide EMRI waveforms for binaries with aligned spins \cite{AlbeETC23}.

Despite this rich literature, it remains a practical challenge to implement an adiabatic model that is efficient, accurate, and accessible across a majority of the parameter space, especially for arbitrary values of the massive black hole spin. Even when neglecting eccentricity, the inclusion of spin presents a number of issues that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced waveform codes that incorporate all possible orbital effects.

@@

\begin{acknowledgements}
This research was supported by an appointment to the NASA Postdoctoral Program at the NASA Goddard Space Flight Center, administered by Oak Ridge Associated Universities under contract with NASA. The author thanks Michael Katz and Lorenzo Speri for extensive discussions and advice on EMRI data analysis, as well as John G.~Baker and Benjamin Leather for their insight and helpful feedback. This work makes use of the Black Hole Perturbation Toolkit. It also uses the \texttt{Eryn}\footnote{\href{https://github.com/mikekatz04/Eryn}{https://github.com/mikekatz04/Eryn}} and \texttt{lisatools}\footnote{\href{https://github.com/mikekatz04/LISAanalysistools}{https://github.com/mikekatz04/LISAanalysistools}} repositories authored by Michael Katz.
\end{acknowledgements}
